Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)

### 2023-09-06 ms/ms.qmd
fixing issues with make

\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the ratio of photosynthesis in an amphistomatous leaf to that of the same leaf with the same total stomatal conductance but with gas exchange blocked through the upper (adaxial) surface, which we term ʻpseudohypostomyʻ. We thus estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).

\textbf{Key results}  Coastal ʻilima leaves benefit 4.58 times more from amphistomy compared to their montane counterparts. Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and that leaves with more similar conductances at each surface benefit more than leaves with few stomata on the upper surface.

>>>>

\textbf{Methods}  We developed a new method to quantify "amphistomy advantage", $\mathrm{AA}$, as the ratio of photosynthesis in an amphistomatous leaf to that of the same leaf with the same total stomatal conductance but with gas exchange blocked through the upper (adaxial) surface, which we term "pseudohypostomy". We thus estimated $\mathrm{AA}$ and related physiological and anatomical traits in 12 populations, six coastal (open, sunny) and six montane (closed, shaded), of the indigenous Hawaiian species ʻilima (*Sida fallax*).

\textbf{Key results}  Coastal ʻilima leaves benefit `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy compared to their montane counterparts. Our evidence was equivocal with respect to two hypotheses – that coastal leaves benefit more because they are thicker and therefore have lower CO$_2$ conductance through the internal airspace, and that leaves with more similar conductances at each surface benefit more than leaves with few stomata on the upper surface.
